\section{Introduction}
\label{sec:intro}

The integration of spectroscopic imaging with deep learning has produced
a class of compositional models that has become the de facto standard
across infrared tissue classification, hyperspectral remote sensing,
Raman imaging, and mass spectrometry imaging. A canonical pipeline
processes a hyperspectral input $X \in \R^{H \times W \times S}$ in two
stages: a per-pixel spectral reduction $f_\thetap: \R^S \to \R^K$
followed by a spatial model $g_\phip: \R^{K \times H \times W} \to
\R^{N_{\mathrm{cls}} \times H \times W}$. End-to-end joint training of
this pipeline is overwhelmingly the default — adopted by foundational
work in FTIR histology~\cite{berisha2019deep}, recent state-of-the-art
spatial-spectral ViTs for prostate tissue
classification~\cite{oleary2026spatial}, and essentially all
hyperspectral remote sensing systems.

\paragraph{An empirical paradox.}
Despite the prevalence of this paradigm, an unexpected and persistent
pattern has emerged across multiple subfields. Frozen pretrained
spectral encoders dramatically outperform end-to-end joint training
on held-out test data. In our motivating tissue classification setting,
a learned-from-scratch linear $f_\thetap$ trained end-to-end achieves
$0.675$ test F1; freezing a pretrained $f_\thetap$ yields $0.84$--$0.95$
test F1. Most strikingly, replacing $f_\thetap$ with \emph{frozen
random weights} yields $0.70$ test F1 --- outperforming the joint
trained model. Recent independent observations
\cite{oleary2026spatial,berisha2019deep} have noted, in different
terms, that aggressive spectral compression to $16$ features has
negligible impact on spatial-spectral neural network performance, and
have concluded that ``tissue classification only needs a small set of
spectral features.''

We argue the opposite. The reason end-to-end joint training fails is
not that spectral information is unimportant. It is that
\emph{end-to-end optimization fails to extract that information}, even
when the spectral signal is rich and discriminative. The problem is in
the training procedure, not the data.

\paragraph{This paper.}
We provide a rigorous explanation for the failure of joint
spectral-spatial training, grounded in the geometry of the loss
landscape and the dynamics of stochastic gradient descent on
compositional architectures with asymmetric capacity. Two main
theorems form our contribution.

\begin{itemize}
    \item \textbf{Theorem~\ref{thm:hessian} (Hessian Capacity Bound).}
          We bound the condition number of the joint Gauss-Newton
          matrix in terms of the ratio of module capacities. Under
          asymmetry $\Cg \gg \Cf$, the loss landscape is geometrically
          ill-conditioned with $\kappa(G) = \Omega(\Cg/\Cf)$. The
          proof combines block matrix analysis (Schur complement) with
          the mean-field FIM scaling of Karakida et
          al.~\cite{karakida2019universal}.
    \item \textbf{Theorem~\ref{thm:twoscale} (Two-Timescale Spurious
          Convergence).} The geometric ill-conditioning induces an
          implicit slow-fast dynamical system. The high-capacity
          spatial module rapidly fits the data using shortcuts in the
          (initially random) feature map, driving the cross-entropy
          residual to zero exponentially fast. The slow spectral
          module is gradient-starved before it can specialize. The
          system converges to a stationary point at which the spectral
          parameters have not moved appreciably from initialization.
          The proof combines Borkar's two-timescale stochastic
          approximation~\cite{borkar2008stochastic} with Pezeshki et
          al.'s gradient starvation
          framework~\cite{pezeshki2021gradient}.
\end{itemize}

\paragraph{A real-time diagnostic.}
The pathology described by Theorems~\ref{thm:hessian} and
\ref{thm:twoscale} is invisible to standard loss curves --- training
and validation loss both decay smoothly while the spectral module is
silently starved. To make the pathology observable in practice we
introduce the \emph{Effective Gradient Ratio} (EGR), a scalar
diagnostic computable inside any training loop:
\[
    \EGR(t) = \frac{\norm{\nabla_\thetap \loss(\thetap_t, \phip_t)}}
                   {\norm{\nabla_\phip \loss(\thetap_t, \phip_t)}}.
\]
Proposition~\ref{prop:egr_monotonic} establishes that EGR decays along
the slow-fast manifold predicted by Theorem~\ref{thm:twoscale}.
Empirically, the depth of the EGR collapse scales monotonically with
capacity ratio, providing a quantitative, training-time observable for
gradient starvation.

\paragraph{Prescription.}
The theory yields a direct prescription: freeze the spectral module,
or inject a fixed residual baseline. Freezing eliminates the timescale
gap by removing $\thetap$ from the optimization entirely. A residual
baseline ($Z = f_\thetap(X) + h(X)$ for a fixed $h$, e.g.\ PCA) is a
softer intervention that breaks the shortcut without precluding spectral
fine-tuning. We discuss both in Section~\ref{sec:discussion}.

\paragraph{Universality.}
Although our motivating application is infrared tissue classification,
the mechanism we identify applies to any compositional architecture
with capacity asymmetry between sequential modules. We verify in
Section~\ref{sec:experiments} that the same pattern manifests across
linear, CNN, and Vision Transformer spatial models on synthetic
hyperspectral problems, indicating the phenomenon is architecture-
agnostic. We expect the same mechanism to govern frozen-encoder
practices in Raman spectroscopy, mass spectrometry imaging, NIR
analysis, and beyond.

\paragraph{Relation to prior work.}
The closest prior work is in multimodal learning, where Wang et
al.~\cite{wang2020what} and Peng et al.~\cite{peng2022balanced}
identified ``modality competition'' and proposed gradient blending and
modulation interventions. Our setting is unimodal (a single
hyperspectral input) but the mechanism is structurally analogous: the
spatial and spectral subspaces of the input act as competing
``modalities,'' with the spatial module monopolizing the gradient
budget. We discuss this connection further in
Section~\ref{sec:related}.

\paragraph{Outline.}
Section~\ref{sec:related} reviews related work. Section~\ref{sec:setup}
establishes notation. Section~\ref{sec:thm1} proves
Theorem~\ref{thm:hessian}. Section~\ref{sec:thm2} proves
Theorem~\ref{thm:twoscale}. Section~\ref{sec:egr} formalizes the EGR
diagnostic. Section~\ref{sec:experiments} validates both theorems and
the diagnostic on synthetic spectral-spatial problems with linear, CNN,
and ViT spatial models. Section~\ref{sec:real} briefly connects the
synthetic findings to real FTIR/QCL tissue classification data.
Section~\ref{sec:discussion} discusses prescriptions, limitations, and
open problems.
