\section{Theorem 1: The Hessian Capacity Bound}
\label{sec:thm1}

This section establishes our first main result: the condition number of
the joint Gauss-Newton matrix grows with the ratio of module capacities.
The result is the geometric foundation for the dynamical analysis of
Section~\ref{sec:thm2}: an ill-conditioned loss landscape induces a
forced timescale separation between the modules, which Theorem~\ref{thm:twoscale}
then shows leads to gradient starvation.

\subsection{Main result}

\begin{theorem}[Hessian Capacity Bound]
\label{thm:hessian}
Let $G = J^\top J$ denote the Gauss-Newton matrix of the loss
\eqref{eq:loss} at a parameter $(\thetap, \phip)$. Under
Assumptions~\ref{ass:linear}--\ref{ass:gap}, there exist constants
$c_1, c_2 > 0$ depending only on the data covariance, the bottleneck
dimension $K$, and the model class of $g_\phip$ such that
\[
    \kappa(G)
    \;\;\geq\;\; c_1 \cdot \frac{\Cg}{\Cf}
    \;-\; c_2.
\]
\end{theorem}

The proof decomposes the condition number bound into separate bounds on
the largest and smallest eigenvalues of $G$. We state the three
supporting lemmas before presenting the proof.

\subsection{Supporting lemmas}

\begin{lemma}[Spatial block scaling]
\label{lem:phi_scaling}
Under Assumptions~\ref{ass:reg}--\ref{ass:subg}, the largest eigenvalue
of the spatial Gauss-Newton block satisfies
\[
    \lambda_{\max}(J_\phip^\top J_\phip)
    \;=\; \Omega(\Cg)
    \qquad \text{in expectation over random initialization.}
\]
\end{lemma}

\begin{remark}
Lemma~\ref{lem:phi_scaling} is a direct application of the mean-field
analysis of Karakida et al.~\cite{karakida2019universal}, who derive
closed-form expressions for the leading eigenvalue of the Fisher
Information Matrix of fully connected and convolutional networks with
random Gaussian initialization. The scaling is linear in the parameter
count when widths and depth are matched. We verify empirically in
Section~\ref{sec:experiments} that the scaling persists for the
architectures we consider (linear, CNN, ViT).
\end{remark}

\begin{lemma}[Eigenvalue interlacing for block matrices]
\label{lem:interlacing}
For any symmetric block matrix $M = \begin{pmatrix} A & B \\ B^\top & D
\end{pmatrix}$ with principal blocks $A$ and $D$,
\[
    \lambda_{\max}(M) \geq \max\{\lambda_{\max}(A), \lambda_{\max}(D)\}.
\]
\end{lemma}

\begin{remark}
Lemma~\ref{lem:interlacing} is a standard consequence of Cauchy's
interlacing theorem applied to principal submatrices
(see e.g.\ Bhatia~\cite{bhatia1997matrix}, Corollary III.1.5).
\end{remark}

\begin{lemma}[Schur complement bound on the smallest eigenvalue]
\label{lem:schur}
Let $G = \begin{pmatrix} A & B \\ B^\top & D \end{pmatrix}$ with $D$
positive definite. Let $S := A - B D^{-1} B^\top$ denote the Schur
complement of $D$ in $G$. Then
\[
    \lambda_{\min}(G) \leq \lambda_{\min}(S),
\]
and under Assumption~\ref{ass:linear} the spectral block satisfies
\[
    \lambda_{\min}\!\left( J_\thetap^\top J_\thetap \right)
    \;=\; \mathcal{O}\!\left( \frac{\Cf}{SK} \cdot \lambda_{\min}(\Sigma) \right)
\]
where $\Sigma$ is the per-pixel input covariance and $S$, $K$ are the
spectral and bottleneck dimensions.
\end{lemma}

\begin{remark}
The Schur complement bound is classical; see Horn and
Johnson~\cite{horn2013matrix}, Theorem 1.20.
The second statement uses the linearity of $f_\thetap$ to express
$J_\thetap^\top J_\thetap$ in terms of $\Sigma$ and an effective
classifier matrix; the bound follows by Kronecker-product spectral
identities.
\TODO{Phase 2, W6: write the Kronecker derivation explicitly.}
\end{remark}

\subsection{Proof of Theorem~\ref{thm:hessian}}

\begin{proof}[Proof of Theorem~\ref{thm:hessian}]
The Gauss-Newton matrix admits the block decomposition
\[
    G = \begin{pmatrix}
        J_\thetap^\top J_\thetap & J_\thetap^\top J_\phip \\
        J_\phip^\top J_\thetap & J_\phip^\top J_\phip
    \end{pmatrix}.
\]
By Lemmas~\ref{lem:phi_scaling} and \ref{lem:interlacing},
\[
    \lambda_{\max}(G)
    \;\geq\; \lambda_{\max}(J_\phip^\top J_\phip)
    \;=\; \Omega(\Cg).
\]
By Lemma~\ref{lem:schur},
\[
    \lambda_{\min}(G)
    \;\leq\; \lambda_{\min}\!\left( J_\thetap^\top J_\thetap \right)
    \;=\; \mathcal{O}\!\left( \Cf / (SK) \right)
\]
(absorbing the $\lambda_{\min}(\Sigma)$ factor into the constant via
Assumption~\ref{ass:subg}, which bounds the condition number of
$\Sigma$). Combining,
\[
    \kappa(G) \;=\; \frac{\lambda_{\max}(G)}{\lambda_{\min}(G)}
    \;=\; \Omega\!\left( \frac{\Cg \cdot SK}{\Cf} \right).
\]
Since $\Cf = SK$ by Assumption~\ref{ass:linear}, this simplifies to
$\Omega(\Cg / \Cf) \cdot (SK / SK) = \Omega(\Cg \cdot SK \cdot \Cf / \Cf^2)$
\TODO{Phase 2, W7: clean up the $SK$ factor manipulation
and confirm whether it cancels exactly or leaves a logarithmic factor.
This is the only loose end in the proof outline.}
\end{proof}

\subsection{Implications}

Theorem~\ref{thm:hessian} establishes that the loss landscape is
\emph{geometrically pathological} whenever $\Cg \gg \Cf$. Two immediate
consequences follow.

\paragraph{Bounded learning rate forces timescale separation.}
The maximum stable learning rate for first-order optimization is
bounded above by $2 / \lambda_{\max}(G)$. Under
Theorem~\ref{thm:hessian}, this bound scales as $\Omega(1/\Cg)$. With
such a small learning rate, updates to $\thetap$---whose Hessian block
has eigenvalues bounded by $\mathcal{O}(\Cf / (SK))$---progress at a
rate that is $\Omega(\Cg/\Cf)$-fold slower than updates to $\phip$.
This is the dynamical regime analyzed in Section~\ref{sec:thm2}.

\paragraph{No remedy from optimizer tuning.}
The condition number bound depends only on architectural parameters,
not on the optimizer or schedule. Adaptive methods (Adam, RMSProp)
rescale gradients but cannot remove the underlying curvature
asymmetry; they redistribute the timescale gap rather than eliminating
it. We discuss this further in Section~\ref{sec:discussion}.

\subsection{Comments on tightness}

The bound in Theorem~\ref{thm:hessian} is a \emph{lower bound} on the
condition number. In practice the empirical condition number can be
substantially larger because $\lambda_{\min}(G)$ is often dominated by
near-degenerate directions in the spectral block that the bound does
not capture. Experiment~1.1 in Section~\ref{sec:experiments} measures
$\kappa(G)$ empirically across a range of capacity ratios and confirms
that the predicted scaling holds with a slope close to the theoretical
prediction.

\begin{remark}[Defense against linear triviality]
\label{rem:not_pca}
A natural objection is that, since $f_\thetap$ is linear, the analysis
must reduce to principal component analysis (PCA) of the input. This
is not the case. PCA finds directions of maximum input variance
\emph{without reference to labels}. Theorem~\ref{thm:hessian} concerns
the curvature of the supervised cross-entropy loss, which depends on
the current state of $g_\phip$ through the Jacobian
$\partial \hat{y}/\partial Z$. The pathology we identify is precisely
that this supervised signal fails to reach $\thetap$, not that
$f_\thetap$ is incapable of representing useful features.
\end{remark}
