\section{Theorem 2: Two-Timescale Spurious Convergence}
\label{sec:thm2}

Theorem~\ref{thm:hessian} establishes that the loss landscape is
geometrically ill-conditioned under capacity asymmetry. This section
translates that geometric pathology into a dynamical one: under joint
SGD, the spectral parameters $\thetap$ are gradient-starved by the
spatial parameters $\phip$ and the system converges to a stationary
point at which the spectral module has not specialized to any
meaningful feature representation. The proof combines two-timescale
stochastic approximation (Borkar~\cite{borkar2008stochastic}) with the
cross-entropy gradient starvation framework of Pezeshki et
al.~\cite{pezeshki2021gradient}.

\subsection{Continuous-time limit and timescale separation}

We analyze the continuous-time limit of joint SGD, in which the
parameters $(\thetap_t, \phip_t)$ evolve according to the gradient
flow
\begin{equation}
\label{eq:gf}
    \dot{\thetap}(t) = -\nabla_\thetap \loss(\thetap, \phip),
    \qquad
    \dot{\phip}(t) = -\nabla_\phip \loss(\thetap, \phip).
\end{equation}
This limit is standard under the Robbins-Monro step-size schedule and
captures the leading-order dynamics for sufficiently small learning
rates. The presence of stochastic gradient noise modifies the constants
but not the qualitative behavior~\cite{borkar2008stochastic}.

\begin{lemma}[Effective timescale]
\label{lem:timescale}
Define the effective timescale ratio
\[
    \varepsilon \;:=\; \frac{\lambda_{\min}(H_{\thetap\thetap})}
                           {\lambda_{\max}(H_{\phip\phip})}.
\]
Under the conditions of Theorem~\ref{thm:hessian},
$\varepsilon = \mathcal{O}(\Cf / \Cg)$ and hence $\varepsilon \to 0$
in the regime of Assumption~\ref{ass:gap}.
\end{lemma}

\begin{proof}
Immediate from Lemmas~\ref{lem:phi_scaling} and \ref{lem:schur}.
\end{proof}

After rescaling, the gradient flow~\eqref{eq:gf} takes the form of a
singularly perturbed system
\begin{equation}
\label{eq:slowfast}
    \dot{\phip}(t) = -\nabla_\phip \loss(\thetap, \phip),
    \qquad
    \dot{\thetap}(t) = -\varepsilon \cdot \nabla_\thetap^{\mathrm{rescaled}} \loss(\thetap, \phip),
\end{equation}
where $\nabla_\thetap^{\mathrm{rescaled}}$ denotes the gradient
in the geometry induced by the spectral Hessian block.

\subsection{Fast equilibrium manifold}

Under~\eqref{eq:slowfast}, $\phip$ moves on a fast timescale and
$\thetap$ on a slow one. The classical singular perturbation theorem
of Tikhonov (see~\cite{borkar2008stochastic}, Chapter 6, and
\cite{heusel2017gans} for the machine-learning adaptation) implies that
$\phip$ converges rapidly to a quasi-equilibrium manifold parametrized
by $\thetap$.

\begin{lemma}[Fast equilibrium]
\label{lem:fast_eq}
Under~\eqref{eq:slowfast} and the regularity Assumption~\ref{ass:reg},
there exists a smooth manifold
$\phip^\star: \R^{\Cf} \to \R^{\Cg}$ such that for any fixed
$\thetap$, the fast dynamics $\dot{\phip} = -\nabla_\phip
\loss(\thetap, \phip)$ converge to $\phip^\star(\thetap)$ in time
$\mathcal{O}(1)$.
\end{lemma}

\begin{remark}
$\phip^\star(\thetap)$ is the configuration of the spatial model that
best classifies the data given the current (and from the spatial
model's view, frozen) spectral features. Importantly, the cross-entropy
saturation property described below operates \emph{on this manifold},
not in an asymptotic limit.
\end{remark}

\subsection{Cross-entropy saturation and residual decay}

The crucial dynamical step is that, along the fast manifold
$\phip^\star(\thetap)$, the cross-entropy residual decays exponentially
fast. This is the gradient starvation phenomenon of
Pezeshki et al.~\cite{pezeshki2021gradient} specialized to our
compositional setting.

\begin{lemma}[Residual decay along the fast manifold]
\label{lem:residual}
Let $r(t) := \mathrm{softmax}(F_{\thetap, \phip^\star(\thetap)}(X))
- y_{\mathrm{onehot}}$ denote the per-pixel cross-entropy residual
under the fast equilibrium. Assume the spatial model has sufficient
capacity to interpolate the training data \emph{using any feature map
that contains a sufficient spatial signal}---a condition we make
precise in the proof. Then
\[
    \norm{r(t)}^2 \;\leq\; C \cdot e^{-\mu t}
\]
for constants $C, \mu > 0$ depending on the data and architecture.
\end{lemma}

\begin{remark}
The proof of Lemma~\ref{lem:residual} closely follows the NTK analysis
of \cite{pezeshki2021gradient} for cross-entropy losses. The key
difference in our setting is that the dominant features along which the
residual decays are spatially-encoded rather than learned-spectrally:
the spatial model, having enough capacity to exploit positional and
contextual structure in the input, achieves low training loss without
requiring the spectral module to specialize.
\TODO{Phase 3, W10: write the NTK-based residual decay proof in full,
adapting Pezeshki Theorem 3.1 to the compositional setting. Verify the
spatial-signal sufficiency condition holds in our synthetic experiments.}
\end{remark}

\subsection{Main result}

\begin{theorem}[Two-Timescale Spurious Convergence]
\label{thm:twoscale}
Under Assumptions~\ref{ass:linear}--\ref{ass:gap} and the spatial-signal
sufficiency condition of Lemma~\ref{lem:residual}, the joint gradient
flow~\eqref{eq:gf} converges to a point $(\thetap^\star, \phip^\star)$
satisfying
\[
    \nabla \loss(\thetap^\star, \phip^\star) = 0,
\]
and
\[
    \norm{\thetap^\star - \thetap_0} = \mathcal{O}(\varepsilon),
\]
where $\thetap_0$ is the initialization. In particular, the spectral
module is asymptotically pinned to a small neighborhood of its
initialization, independently of the labels.
\end{theorem}

\begin{proof}[Proof sketch]
By Lemma~\ref{lem:fast_eq}, after time $T_{\mathrm{fast}}$ of order
$\mathcal{O}(1)$, the fast variable $\phip$ has converged to
$\phip^\star(\thetap_0)$. By Lemma~\ref{lem:residual}, along this
manifold the residual $r$ decays exponentially:
$\norm{r(t)} \leq C e^{-\mu t}$.

The slow-variable update~\eqref{eq:slowfast} satisfies
\[
    \norm{\dot{\thetap}(t)} = \varepsilon \cdot \norm{J_\thetap^\top r(t)}
    \leq \varepsilon \cdot \norm{J_\thetap}_{\mathrm{op}} \cdot C \cdot e^{-\mu t}.
\]
Integrating from $T_{\mathrm{fast}}$ to $\infty$,
\[
    \norm{\thetap^\star - \thetap_0}
    \leq \varepsilon \cdot \norm{J_\thetap}_{\mathrm{op}} \cdot C \cdot \frac{e^{-\mu T_{\mathrm{fast}}}}{\mu}
    = \mathcal{O}(\varepsilon).
\]
The fixed-point condition $\nabla \loss = 0$ holds because
both $r \to 0$ (so $\nabla_\thetap \loss \to 0$) and
$\phip$ is at the fast equilibrium ($\nabla_\phip \loss = 0$ on the
manifold).
\end{proof}

\subsection{Interpretation}

Theorem~\ref{thm:twoscale} formalizes the central claim of the paper.
The point $(\thetap^\star, \phip^\star)$ is mathematically a critical
point of the loss but is, in any practical sense, a failure: the
spectral parameters have not departed materially from their
initialization, and the spatial module has interpolated the labels
using whatever spatial signal exists in the (near-random or pretrained)
feature map. Whether the spectral module \emph{could have learned}
useful features under freezing-then-decoupling or other interventions
is precisely what our prescriptive remedy (Section~\ref{sec:discussion})
addresses.

\subsection{Comments on the assumptions}

The spatial-signal sufficiency condition deserves comment. We require
that the spatial model can interpolate the training labels using only
the structure of the feature map $Z$ (e.g., positional cues, contextual
patterns), without requiring $f_\thetap$ to specialize. In practice
this condition is satisfied whenever the spatial model has sufficient
capacity to overfit the training set---which is the norm for modern
overparameterized ViTs and CNNs. We verify the condition empirically
in Experiments~1.2 and 1.3 by showing that joint training reaches near-
zero training loss with a spectral module that has not appreciably
changed from initialization.

\TODO{Phase 3, W10: Formalize the spatial-signal sufficiency condition.
Likely route: bound on Rademacher complexity of $g_\phip \circ Z$ that
guarantees zero-error interpolability without requiring rank growth in
$Z$. Cross-check against \cite{pezeshki2021gradient} Section 3.}
