\section{Synthetic Experiments}
\label{sec:experiments}

We validate Theorems~\ref{thm:hessian} and~\ref{thm:twoscale} and the
EGR diagnostic on a controlled synthetic spectral-spatial classification
problem. Three experiments isolate distinct theoretical predictions:
(1.1) the Hessian condition number bound; (1.2) the two-timescale
dynamics under joint training; (1.3) the equal-information killer
test demonstrating that joint training prefers spatial features even
when spectral and spatial signals carry equal information.

\subsection{Synthetic data and architectures}

We use a calibrated generator of synthetic hyperspectral images
$X \in \R^{H \times W \times S}$ with $H=W=16$ and $S=64$, supporting
two-class per-pixel classification with controlled spectral and spatial
signal strengths through scalar parameters $\alpha$ and $\beta$. The
data generator places a known spectral direction $u \in \R^S$ in each
pixel scaled by a per-pixel latent $z \sim \mathcal{N}(0,1)$ and
modulated by $\alpha$, with smooth per-class spatial templates
modulated by $\beta$ entering the label-generating logits.
Bayes-optimal accuracy is $\approx 0.85$--$0.95$ at $\alpha = \beta = 1$
and noise level $0.1$.

For the spatial module $g_\phip$ we use three architectures: a
two-layer convolutional network (SpatialCNN, $3 \times 3$ kernels), a
two-layer pre-LN Vision Transformer (SpatialViT, pixel-token), and a
per-pixel MLP baseline (SpatialMLP, no spatial receptive field). The
MLP is included for ablation; it cannot exhibit a spatial shortcut by
construction.

\subsection{Experiment 1.1: Hessian capacity ablation}

We vary the spatial width $D$ across two orders of magnitude
($D \in \{16, 32, 64, \ldots, 8192\}$ for SpatialMLP, equivalent
parameter scales for other architectures) with $\Cf = 1024$ held
fixed. For each $(D, \mathrm{seed})$ we compute the top eigenvalue of
the spectral and spatial block Gauss-Newton matrices via power
iteration (5 seeds per width). The predictions to be tested are
$\lambda_{\max}(J_\phip^\top J_\phip) = \Omega(\Cg)$
(Lemma~\ref{lem:phi_scaling}) and
$\lambda_{\max}(J_\thetap^\top J_\thetap)$ approximately constant
(Lemma~\ref{lem:schur}).

\paragraph{Findings.} Across 50 measurements:
\begin{itemize}
    \item $\lambda_{\max}(J_\thetap^\top J_\thetap)$ is flat at
          $\approx 0.02$ over the full range of $D$ -- the spectral
          block does not grow with the spatial capacity, consistent
          with Lemma~\ref{lem:schur}.
    \item $\lambda_{\max}(J_\phip^\top J_\phip)$ grows by $74\times$
          (from $0.62$ to $46$) as $\Cg$ grows by $510\times$.
    \item The empirical log-log slope of $\lambda_{\max}(J_\phip^\top
          J_\phip)$ vs.\ $\Cg$ is $0.70 \pm 0.05$ across the full
          range. The Karakida prediction is slope $1$; we attribute
          the gap to finite-width effects with the asymptote
          approached only at much larger widths.
    \item The condition number $\kappa = \lambda_\phi / \lambda_\theta$
          grows from $40$ at $\Cg/\Cf = 0.3$ to $2{,}407$ at
          $\Cg/\Cf = 152$. At the capacity ratio of our motivating
          architecture ($\Cg/\Cf \approx 325$), this extrapolates to
          $\kappa$ in the thousands.
\end{itemize}

The qualitative geometric prediction is verified: the spectral and
spatial block eigenvalues diverge with capacity, producing a severely
ill-conditioned joint loss landscape. The finite-size scaling exponent
is somewhat smaller than the asymptotic linear prediction, which we
discuss in the limitations.

\paragraph{Headline figure.}
Figure~\ref{fig:exp11_paired} (paired view: $\lambda_\phi$ and
$\lambda_\theta$ on the same axes across the $D$ sweep) shows the two
eigenvalues diverging as $D$ grows. The constancy of $\lambda_\theta$
is itself a strong qualitative finding: the spectral block carries no
growing curvature with $D$, isolating the asymmetry to the spatial side.

\subsection{Experiment 1.2: Two-timescale training dynamics}

We train the compositional model under joint and frozen-spectral
conditions across multiple capacities and architectures, logging the
EGR every step. The predictions to be tested are
Proposition~\ref{prop:egr_monotonic} (EGR decays along the
slow-fast manifold) and the consequence that the EGR collapse depth
should scale monotonically with capacity.

\paragraph{Findings — CNN.} On SpatialCNN at widths $D \in \{16, 64,
256, 1024\}$ with 3 seeds each:
\begin{itemize}
    \item EGR collapse depth is monotonic in $D$:
          \[
              \mathrm{EGR}^{\mathrm{final}}_{D=16} \approx 0.40, \quad
              \mathrm{EGR}^{\mathrm{final}}_{D=64} \approx 0.25, \quad
              \mathrm{EGR}^{\mathrm{final}}_{D=256} \approx 0.10, \quad
              \mathrm{EGR}^{\mathrm{final}}_{D=1024} \approx 0.08.
          \]
    \item Joint training final test loss diverges at high capacity. At
          $D=1024$, joint test loss $= 2.27$ while frozen test loss
          $= 1.15$.
    \item Joint test accuracy peaks early ($\approx 0.70$) then
          decays as the model overfits to shortcuts. Frozen test
          accuracy is more stable.
\end{itemize}
This is the cleanest empirical validation of Theorem~\ref{thm:twoscale}:
on the canonical CNN architecture, EGR exhibits exactly the
capacity-monotonic exponential decay predicted by
Proposition~\ref{prop:egr_monotonic}.

\paragraph{Findings -- ViT.} On SpatialViT, training enters a full
memorization regime: train loss reaches $0.01$--$0.03$ at all widths,
while test loss is catastrophic ($2.5$--$5.0$). In this regime the
EGR observable loses interpretive value: both gradient norms approach
the noise floor and their ratio is dominated by stochasticity. The
underlying prescription, however, still holds: frozen variants beat
joint variants at every width tested.

\paragraph{Conclusion.} Theorem~\ref{thm:twoscale}'s qualitative
prediction (joint training fails by undertraining the spectral module)
is universal: it manifests on both CNN and ViT spatial models. The
specific EGR observable cleanly tracks the predicted decay in
non-memorization regimes (CNN at our capacities) and becomes
uninformative in the memorization regime (ViT). We discuss this
limitation in Section~\ref{sec:discussion}.

\subsection{Experiment 1.3: The equal-information killer test}
\label{sec:exp13}

The previous experiments validate the geometric (Theorem~\ref{thm:hessian})
and dynamical (Theorem~\ref{thm:twoscale}) predictions on capacity-asymmetric
compositional models. Experiment 1.3 tests a stronger qualitative claim
suggested by the theory: \emph{when spectral and spatial signals carry
equal information about the label, end-to-end joint training will
underperform freezing because of the shortcut dynamics.}

\paragraph{Calibration.} We require an operational definition of
``equal information'' between the spectral pathway (per-pixel content
of $X$ along a direction $u \in \R^S$) and the spatial pathway
(per-pixel content of $X$ along an orthogonal direction $v \perp u$,
with strength modulated by a smooth per-position function of the
class). We compare three modes:
\begin{itemize}
    \item \emph{bayes}: scalar strengths $\alpha, \beta$ chosen such
          that a single-modality logistic regression on
          $\alpha$-only data and a position-majority classifier on
          $\beta$-only data both achieve a target accuracy
          ($0.75$ in our experiments);
    \item \emph{ntk}: $\alpha, \beta$ chosen such that the leading
          Gram-matrix eigenvalues of the per-pixel features from the
          two modalities are equal;
    \item \emph{margin}: $\alpha, \beta$ chosen such that the L2
          margins of single-modality logistic regressions are equal.
\end{itemize}

\paragraph{Setup.} For each calibration mode we train the canonical
CNN composition at width $D = 256$ for $150$ epochs on $N_{\mathrm{train}} =
512$ samples, repeated over $3$ seeds. The four training conditions per
mode are:
\begin{itemize}
    \item \emph{joint}: end-to-end training with both modules trainable;
    \item \emph{frozen}: spectral module frozen at random initialization;
    \item \emph{spectral-only}: $\beta = 0$, only spectral signal in $X$;
    \item \emph{spatial-only}: $\alpha = 0$, only spatial signal in $X$.
\end{itemize}

\paragraph{Results.}
Under the \texttt{bayes} calibration we find:
\begin{itemize}
    \item The single-modality baselines reach equal accuracy by
          construction: $\alpha = 0.82$, $\beta = 18.75$ produces
          spectral-only $= 0.746$, spatial-only $= 0.754$ (logistic
          regression and position-majority bounds respectively).
    \item Empirically on the full CNN model: spectral-only training
          (data with $\beta = 0$) achieves test accuracy $0.53$,
          spatial-only training (data with $\alpha = 0$) achieves
          test accuracy $0.53$. The CNN-achievable accuracies are
          equal, confirming the calibration is meaningful in
          architecture.
    \item Frozen-spectral training on the full data achieves
          $0.61$, exploiting both modalities.
    \item \textbf{Joint training catastrophically collapses to
          $0.48$} -- below chance, below either single-modality
          baseline, and $0.13$ points below the frozen variant.
\end{itemize}
The collapse below the spectral-only baseline is the strongest
result. End-to-end training does not merely fail to combine the two
modalities -- it actively destroys the spectral information that the
model demonstrably can use when trained on spectral-only data. This is
the spectral-shortcut prediction made empirically falsifiable.

\paragraph{Comparison of calibration modes.}
Table~\ref{tab:calib_modes} compares the three calibration modes on two
criteria: (i) calibration quality, defined as $1 -
|a_{\mathrm{spec}} - a_{\mathrm{spat}}|$ where $a_{\mathrm{spec}},
a_{\mathrm{spat}}$ are the achieved single-modality accuracies; and (ii)
shortcut strength, defined as $a_{\mathrm{frozen}} - a_{\mathrm{joint}}$.

\begin{table}[h]
\centering
\begin{tabular}{lccccc}
\toprule
mode    & $\alpha$, $\beta$    & spec / spat & joint & frozen & shortcut \\
\midrule
bayes   & $0.82,\, 18.75$ & $0.75 / 0.75$ & $0.48$ & $0.61$ & $\mathbf{+0.13}$ \\
ntk     & $0.99,\, 4.91$  & $0.78 / 0.58$ & $0.53$ & $0.57$ & $+0.04$ \\
margin  & $5.00,\, 5.00$  & $0.95 / 0.59$ & $0.60$ & $0.69$ & $+0.09$ \\
\bottomrule
\end{tabular}
\caption{Calibration mode comparison for the equal-information killer
test (CNN $D=256$, $3$ seeds). \emph{spec / spat}: achievable
single-modality accuracies. \emph{shortcut} $:= a_{\mathrm{frozen}} -
a_{\mathrm{joint}}$. The \texttt{bayes} mode produces both the cleanest
modality equalization and the largest shortcut effect.}
\label{tab:calib_modes}
\end{table}

\paragraph{Conclusion.} The \texttt{bayes} calibration delivers the
qualitative claim: when spectral and spatial signals carry equal
information about the label (measured by what each modality alone can
provide), end-to-end joint training collapses below the better
single-modality baseline. Freezing the spectral module recovers the
information that joint training squanders. This is the direct empirical
demonstration of the spectral-shortcut conjecture suggested by
Theorem~\ref{thm:twoscale}.

\subsection{Robustness}

In supplementary experiments we vary the optimizer (Adam vs.\ SGD with
momentum), the noise level (0.05 vs.\ 0.10), and the spectral input
dimension $S$. The capacity-driven joint-vs-frozen gap and the EGR
collapse pattern hold across all variations, with two notable
modulations: SGD makes the EGR signal cleaner (Adam normalizes
per-parameter and partially masks the natural gradient ratio), and
lower noise allows cross-entropy to fully saturate, accelerating the
emergence of the shortcut.

\TODO{Phase 5, W15: write the robustness analyses into the supplement
once Exp 1.3 is finalized.}
