\section{Discussion}
\label{sec:discussion}

\subsection{Implications for practice}

The theorems and experiments establish a concrete prescription. In any
compositional spectral-spatial pipeline with substantial capacity
asymmetry $\Cg \gg \Cf$, end-to-end joint training is expected to fail
in a specific and predictable way: the spectral module stays near
initialization while the spatial module fits the data using shortcuts
in the unstructured feature map. Three operational responses follow.

\paragraph{Freeze the spectral module.}
The most direct response to Theorem~\ref{thm:twoscale} is to remove
$\thetap$ from the optimization. Without a spectral update, the
timescale separation is eliminated by construction, and $g_\phip$
optimizes a standard supervised learning problem on fixed input
features. If a pretrained spectral encoder is available (from
unsupervised pretraining, self-supervised pretraining, or pre-task
pretraining on the same data), it should be used frozen.

\paragraph{Use a fixed baseline.}
When pretraining is unavailable, a fixed but informative baseline
$h(X)$ (e.g.\ PCA, fixed wavelet basis, random projection) can be
injected as an additive component $Z = f_\thetap(X) + h(X)$. The
fixed component ensures the spatial model receives a stable input even
if $f_\thetap$ drifts; the additive structure allows fine-grained
refinement of $f_\thetap$ without exposing the spatial model to the
moving-target dynamics that produce the shortcut. We have not
empirically tested this prescription in this paper; it follows
directly from the theory and is a natural focus for follow-up work.

\paragraph{Track EGR.}
The EGR diagnostic detects the onset of starvation in real time.
Practitioners should log it alongside training and validation loss.
When $\EGR(t)$ drops below a chosen threshold (we recommend $0.1
\cdot \EGR(0)$), the theoretical analysis predicts that subsequent
$\thetap$ updates are essentially noise: the spectral module has been
absorbed into a spurious stationary point. At this point, freezing
$\thetap$ and continuing to train $\phip$ retains the run's
investment while avoiding the negative consequences of further joint
updates.

\subsection{Limitations}

We make four limitations explicit.

\paragraph{Linear spectral reduction.}
Theorems~\ref{thm:hessian} and~\ref{thm:twoscale} are proven under the
assumption that $f_\thetap$ is linear in $\thetap$. In practice some
pipelines use shallow nonlinear spectral modules (1--2 layer MLPs).
The intuition behind both theorems extends: nonlinear shallow
$f_\thetap$ retains the property that the spectral block carries
substantially less capacity than the spatial block, so the
qualitative timescale separation persists. We verify empirically in
Section~\ref{sec:experiments} that the joint-vs-frozen gap remains
under nonlinear $f_\thetap$. A formal proof for nonlinear $f_\thetap$
is left to future work.

\paragraph{Finite-width scaling exponent.}
The empirical scaling exponent of $\lambda_\phi$ vs.\ $\Cg$ is $\sim
0.7$ at the widths we tested, while the asymptotic Karakida prediction
is $1.0$. We expect the asymptote to be approached at larger widths.
Two consequences for the paper: the Theorem~\ref{thm:hessian} bound
should be read as a lower bound that may be loose at finite scale, and
the empirical predictions should be interpreted as qualitative
(monotonic growth) rather than tight (constant proportionality).

\paragraph{Memorization regime.}
For sufficiently overparametrized spatial models (ViT in our
experiments), joint training enters a memorization regime in which
train loss falls to near zero and both gradient norms approach the
noise floor. In this regime the EGR observable becomes uninformative,
even though the underlying prescription (freeze) continues to apply.
Practitioners running highly overparametrized models should observe
the symptoms (test loss divergence, train loss saturation) rather than
relying on EGR alone.

\paragraph{Single domain validation.}
Our real-data validation is restricted to FTIR/QCL tissue classification
in breast cancer. We expect the same mechanism to operate in
hyperspectral remote sensing, Raman tissue classification, and mass
spectrometry imaging, but those validations are future work.

\subsection{Defense against the "linear triviality" critique}

A natural reviewer objection is that, because $f_\thetap$ is linear,
the analysis must reduce to principal component analysis. We address
this in Remark~\ref{rem:not_pca} and emphasize again: PCA finds
directions of maximum input variance independently of labels.
Theorem~\ref{thm:hessian} concerns curvature of the supervised
cross-entropy loss, which depends on the current state of $g_\phip$
through the Jacobian $\partial \hat{y}/\partial Z$. The pathology we
identify is that this supervised signal fails to reach $\thetap$. A
PCA-decomposed $f_\thetap$ would not exhibit the joint-vs-frozen gap
we observe: indeed, our experiments include a frozen-PCA baseline that
performs well, exactly because PCA decomposition does not require the
supervised signal to flow through it.

\subsection{The Spatial Dominance Conjecture: an open problem}

A stronger statement than Theorem~\ref{thm:twoscale} is suggested by
our experiments but not proven: \emph{when spectral and spatial
features carry equal information about the label, joint training
exhibits an implicit bias toward the spatial pathway}. Pezeshki et
al.~\cite{pezeshki2021gradient} prove a related but distinct result in
the NTK regime; Huang et al.~\cite{huang2022modality} prove a related
result for explicit dual-modality networks but not for compositional
single-modality pipelines. A proof of the spatial-dominance conjecture
in our setting would require formalizing ``equal information'' in an
operationally meaningful way (Shannon mutual information is bijection-
invariant and therefore not the right primitive for gradient dynamics)
and combining the NTK eigenvalue gap with our compositional Hessian
analysis. We leave this as an open problem.

\subsection{Beyond spectroscopy: where else this matters}

Although our motivating application is infrared tissue classification,
the mechanism we identify applies to any compositional architecture
with capacity asymmetry between sequential modules. We highlight three
areas where the same pathology likely operates.

\paragraph{Hyperspectral remote sensing.}
The dominant paradigm for hyperspectral image classification is a
shallow spectral reduction (PCA, autoencoder, or 1D conv) feeding into
a deep spatial classifier (3D CNN, ViT). The capacity asymmetry is
typically large. Joint training is the de facto standard. Our analysis
predicts the same shortcut emergence; whether the field has noticed it
under the guise of ``classification works without much spectral
processing'' is a question worth examining.

\paragraph{Multimodal foundation models.}
Vision-language and audio-language models compose a sequence of
modality-specific encoders into a shared representation space. The
modality with the larger encoder typically dominates the shared
representation. While this has been documented empirically (Wang et
al.~\cite{wang2020what}, Peng et al.~\cite{peng2022balanced}), the
mechanism is the same as ours: capacity-induced timescale separation
combined with cross-entropy saturation. Our framework extends their
empirical observations into the single-input compositional setting.

\paragraph{Video models with frozen image encoders.}
A standard practice in video understanding is to freeze a pretrained
image encoder and train a temporal model on top. This is widely
treated as ``transfer learning,'' but the same mechanism we describe
predicts that joint training would fail in a specific way -- the
image encoder would drift, the temporal model would absorb the
shortcut, generalization would suffer. The empirical literature
strongly supports the freezing practice; our analysis provides a
mechanistic reason.

\subsection{Concluding remarks}

We have provided the first rigorous explanation for the failure of
end-to-end joint training in compositional spectral-spatial deep
learning. The phenomenon is geometric (Theorem~\ref{thm:hessian}) and
dynamical (Theorem~\ref{thm:twoscale}), is detectable in real time via
the EGR diagnostic, and has a direct prescriptive remedy (freezing or
fixed-baseline injection). We expect the framework to extend beyond
spectroscopy to any compositional architecture with capacity asymmetry,
and to provide a principled foundation for the empirical practice of
frozen feature extractors that the field has converged on by
necessity rather than design.
