\section{Supplementary Material}
\label{sec:supplement}

This supplement provides the full proofs of Theorems~\ref{thm:hessian}
and~\ref{thm:twoscale}, additional experiments referenced in the main
text, and robustness checks.

\subsection{Full proof of Theorem~\ref{thm:hessian}}
\label{sec:proof_thm1}

We restate Theorem~\ref{thm:hessian} for reference:

\begin{quote}
\emph{Under Assumptions \ref{ass:linear}--\ref{ass:gap}, the condition
number of the joint Gauss-Newton matrix satisfies
$\kappa(G) \geq c_1 \cdot (\Cg/\Cf) - c_2$ for some
$c_1, c_2 > 0$ depending only on the data covariance, the bottleneck
dimension $K$, and the model class of $g_\phip$.}
\end{quote}

The proof is structured as three lemmas: (i) the spatial block's top
eigenvalue scales with $\Cg$ (Lemma~\ref{lem:phi_scaling}); (ii)
Cauchy interlacing transfers this scaling to the full Hessian
(Lemma~\ref{lem:interlacing}); (iii) a Schur complement bound caps
the minimum eigenvalue by a $\Cf$-dependent quantity
(Lemma~\ref{lem:schur}). We prove each in turn, then combine.

\subsubsection{Proof of Lemma~\ref{lem:phi_scaling}}

We show $\lambda_{\max}(J_\phip^\top J_\phip) = \Omega(\Cg)$ in
expectation over random Gaussian initialization, under
Assumptions~\ref{ass:reg} (twice-differentiable, $L$-Lipschitz
gradients) and \ref{ass:subg} (sub-Gaussian inputs with bounded
covariance condition number).

\paragraph{Step 1: Bounded per-pixel residual at initialization.}
Under Kaiming initialization, each pre-activation in the spatial
module $g_\phip$ has $\mathcal{O}(1)$ variance. The output logits
$\hat{y} = g_\phip(f_\thetap(X))$ are therefore distributed as
$\hat{y} \sim \mathcal{N}(0, \sigma_y^2 I_{N_{cls}})$ at initialization
in the mean-field limit, with $\sigma_y^2 = \Theta(1)$ independent
of $\Cg$. The softmax probabilities $p = \mathrm{softmax}(\hat{y})$
are concentrated near uniform $1/N_{cls}$, and the per-pixel residual
$r = p - y_{onehot}$ satisfies
\[
    \E\!\left[ \norm{r}^2 \right]
    \;=\; \frac{N_{cls}-1}{N_{cls}} \;+\; \mathcal{O}(\sigma_y^2)
    \;=\; \Theta(1).
\]
This bound is independent of $\Cg$.

\paragraph{Step 2: Trace of the Fisher Information block.}
Karakida et al.~\cite{karakida2019universal}, Theorem~3.1, give that
for a deep network at random Kaiming initialization, the expected
trace of the per-layer FIM block scales linearly with the layer's
parameter count:
\[
    \E\!\left[ \mathrm{tr}(F_\phip) \right]
    \;=\; \sum_{\ell=1}^{L} q_\ell \cdot \nu_\ell \cdot M_\ell,
\]
where $M_\ell$ is the parameter count of layer $\ell$, $q_\ell$ is
the squared activation magnitude at layer $\ell$, and $\nu_\ell$ is
the squared backpropagated signal at layer $\ell$. Under Kaiming
init, $q_\ell, \nu_\ell = \Theta(1)$ uniformly in $\ell$, so
\[
    \E\!\left[ \mathrm{tr}(F_\phip) \right]
    \;=\; \Theta\!\left( \sum_\ell M_\ell \right)
    \;=\; \Theta(\Cg).
\]

\paragraph{Step 3: From Fisher to Gauss-Newton.}
For softmax cross-entropy, the FIM and Gauss-Newton matrix differ by
a factor depending on the prediction distribution:
\[
    F_\phip \;=\; \E_X\!\left[ J_\phip^\top \cdot \mathrm{diag}(p)(I - 1 p^\top) \cdot J_\phip \right].
\]
At random initialization $p \to (1/N_{cls}) \mathbf{1}$, so the
inner matrix has eigenvalues bounded by $1/N_{cls}$. Therefore
\[
    \mathrm{tr}(F_\phip)
    \;\geq\; \frac{1}{N_{cls}^2} \cdot \mathrm{tr}(J_\phip^\top J_\phip)
\]
and combining with Step~2:
\[
    \E\!\left[ \mathrm{tr}(J_\phip^\top J_\phip) \right]
    \;=\; \Omega(\Cg).
\]

\paragraph{Step 4: Top eigenvalue from trace via participation ratio.}
The top eigenvalue is bounded from below by the trace divided by the
effective rank:
\[
    \lambda_{\max}(J_\phip^\top J_\phip)
    \;\geq\; \frac{\mathrm{tr}(J_\phip^\top J_\phip)}{\rho(J_\phip^\top J_\phip)},
\]
where $\rho$ is the participation ratio (effective rank under the
$L^2$ measure). Karakida et al.~\cite{karakida2019universal},
Theorem~4.2, show that under mean-field assumptions, the
participation ratio of the FIM is $\Theta(1)$ uniformly in network
width. Combining with Step~3:
\[
    \lambda_{\max}(J_\phip^\top J_\phip)
    \;=\; \Omega(\Cg).
\]

\paragraph{Step 5: Concentration in the wide-network limit.}
Karakida et al.~\cite{karakida2019universal}, Theorem~5.1, establish
that in the mean-field limit, the top eigenvalue of the FIM
concentrates around its expectation with high probability. The
deterministic bound therefore holds with probability $1 - o(1)$ as
$\Cg \to \infty$.

This completes the proof of Lemma~\ref{lem:phi_scaling}. \hfill$\square$

\begin{remark}[Finite-width scaling]
Karakida's bound is asymptotic in network width. At the finite widths
explored in Experiment~1.1 ($D \in \{16, \ldots, 8192\}$), we observe
an empirical scaling exponent of $\approx 0.7$ rather than the
asymptotic $1$. This sub-asymptotic behaviour is typical in NTK and
mean-field analyses: the participation-ratio bound in Step~4 is tight
only in the strict width limit. The qualitative scaling --- monotonic
linear growth of $\lambda_\phi$ with $C_g$ --- is preserved at all
widths we test.
\end{remark}

\subsubsection{Proof of Lemma~\ref{lem:interlacing}}

The claim that $\lambda_{\max}(M) \geq \max\{\lambda_{\max}(A),
\lambda_{\max}(D)\}$ for a symmetric block matrix
$M = \begin{pmatrix} A & B \\ B^\top & D \end{pmatrix}$ is a direct
application of Cauchy's interlacing theorem to principal submatrices.
See Bhatia~\cite{bhatia1997matrix}, Corollary~III.1.5.

For the eigenvector $v$ achieving $\lambda_{\max}(A) = v^\top A v$,
the extension $\tilde{v} = \begin{pmatrix} v \\ 0 \end{pmatrix}$ has
$\tilde{v}^\top M \tilde{v} = v^\top A v$, so
$\lambda_{\max}(M) \geq \tilde{v}^\top M \tilde{v} / \norm{\tilde{v}}^2
= \lambda_{\max}(A)$. The same argument for $D$ completes the
proof. \hfill$\square$

\subsubsection{Proof of Lemma~\ref{lem:schur}}
\TODO{Phase 2 W7. Write the Schur complement bound for $\lambda_{\min}$
using the Kronecker product expansion of $J_\thetap^\top J_\thetap = \Sigma \otimes A^\top A$, with $A$ the effective downstream classifier. Cite Horn-Johnson Theorem 1.20 for the Schur step and Bhatia for the Kronecker spectral identity. Approximate length: 1.5 pages.}

\subsubsection{Combining the three lemmas}
\TODO{Phase 2 W7. Combine the bounds from Steps 1-5 of
Lemma~\ref{lem:phi_scaling}, Lemma~\ref{lem:interlacing} (transfer to
full Hessian), and Lemma~\ref{lem:schur} (smallest eigenvalue). Clean
up the $S \cdot K$ factor, noted as outstanding in the main text
proof outline. Approximate length: 1 page.}

\subsection{Full proof of Theorem~\ref{thm:twoscale}}
\label{sec:proof_thm2}

\TODO{Phase 3, weeks 8-10. Combine Borkar singular perturbation with
Pezeshki residual decay. Proof skeleton: (i) ODE limits of SGD,
(ii) timescale gap from Lemma 5.1, (iii) Tikhonov fast-manifold
convergence, (iv) NTK residual decay along fast manifold,
(v) slow-variable bound by Gronwall.}

\subsection{Empirical robustness}

We here report robustness experiments referenced in the main text.

\subsubsection{Optimizer ablation (SGD vs Adam)}
\TODO{Phase 5 follow-up. Show EGR collapse pattern in both regimes;
note Adam masks the natural gradient ratio.}

\subsubsection{Loss-variant ablation}
\TODO{Phase 5 follow-up. Cross-entropy, focal loss, label smoothing.}

\subsubsection{Regularizer ablation}
\TODO{Phase 5 follow-up. Weight decay, dropout, batch norm.
Pezeshki et al.~\cite{pezeshki2021gradient}, App.~B, already show
gradient starvation is robust to these; we reproduce briefly in our
setting.}

\subsubsection{Smallest capacity ratio where the pathology disappears}
\TODO{Phase 5 follow-up. Sweep $\Cg/\Cf$ from 1 to 1000, identify
the threshold below which joint training reaches frozen
performance.}
